---
title: "2nd Assignment of 'Reproducible Research'" 
subtitle: "Impact of Storm Events on Health and Economy"
author: "Hauer-Glocke"
date: "18 Juli 2017"
output: html_document
fig_caption: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)

```

#Synopsis - Overview Data Analysis

> Overview in less then 10 sentences

#Data Processing

> describes how the data were loaded into R and processed for analysis

```{r, echo = TRUE, message = FALSE}
#Loading of Data from URL
library(readr)
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",temp)
con <- bzfile(temp)
df <- read_csv(con)
unlink(temp)
rm(con, temp)
```

```{r, echo = TRUE, message = FALSE}
#Identification of relevant Variables
library(dplyr)
df1 <- df %>%
        #Selection of relevant variables (Date, Event, Healt Impact and Economic Impact)
        select(BGN_DATE ,EVTYPE,FATALITIES,INJURIES,PROPDMG, CROPDMG, PROPDMGEXP, CROPDMGEXP) %>%
        mutate(cases_harm = FATALITIES + INJURIES) %>%
        mutate(economic_damage = PROPDMG + CROPDMG)
```


#Results

> main results are presented?

##1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

The Harm on the Population is measured via the Number of Fatality and Injury cases, as indicated "FATALITIES" and "INJURIES". Further I use the sum to indicate the "Harm to Population Health".


```{r, echo = TRUE, message = FALSE}
#Build Dataset for harmed Population
library(lubridate)

df1$BGN_DATE <- as.Date(df1$BGN_DATE, "%d/%m/%Y %H:%M:%S")
df_harm1 <- df1 %>%
        filter(!(is.na(BGN_DATE))) %>%
        group_by(BGN_DATE) %>%
        summarise(harm_by_date = sum(cases_harm))
```

```{r fig1, echo = TRUE, fig.cap= "This graph shows a timeline of harmful events in the US. The harmful cases are aggregated on the day."}
plot(df_harm1, 
     type = "l",
     main = "Harmful events to the population",
     xlab = "Timeline of events",
     ylab = "Number of harmed people (injured or fatal)")

```

```{r, echo = TRUE, message = FALSE}
#What events harm the population most?
df_harm2 <- df1 %>%
        group_by(EVTYPE) %>%
        summarise(total_harm = sum(cases_harm)) %>%
        arrange(desc(total_harm))
head(df_harm2, 5)

```

##2. Across the United States, which types of events have the greatest economic consequences?

Economic Consequences are defined as Properitory and Crop Damages, this is indicated with the variables "PROPDMG" and "CROPDMG". Further I use the sum to indicate the "Economic Consequences".

```{r, echo = TRUE, message = FALSE}
#Developement of Economic Consequences over all Event Types
df_dmg <- df1 %>%
        group_by(EVTYPE) %>%
        summarise(economic_damage_total = sum(economic_damage)) %>%
        arrange(desc(economic_damage_total))

#Event Types with greatest Economic Consequences
head(df_dmg, 5)

```

```{r fig2, echo = TRUE, fig.cap = "Here I can add a long which describes the above viewed figure. I guess even when the text is long, it should not occur a problem here?"}




```

